Modeling Lake Trophic State: A Data Mining Approach

### 2014-09-19 inst/doc/manuscript.Rmd
Updating analysis, new function added, and updated data

---
title: 'Modeling Lake Trophic State: A Data Mining Approach'
author: 
- name: Jeffrey W. Hollister
  affilnum: 1
  email: hollister.jeff@epa.gov
- name: W. Bryan Milstead
  affilnum: 1
- name: Betty J. Kreakie
  affilnum: 1
affiliation:
- affilnum: 1
  affil: US Environmental Protection Agency, Office of Research and Development, National     Health and Environmental Effects Research Laboratory, Atlantic Ecology Division, 27 Tarzwell Drive  Narragansett, RI, 02882, USA
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    template: components/manuscript.latex
  html_document: null
  word_document: null
fontsize: 11pt
capsize: normalsize
csl: components/ecology.csl
documentclass: article
spacing: doublespacing
linenumbers: no
bibliography: components/LakeTrophicModelling.bib

---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Modeling Lake Trophic State: A Data Mining Approach}
-->

@@

```{r analysis , include=FALSE, echo=FALSE}
#All analysis in here, that way all bits of the paper have access to the final objects
#Place tables and figures and numerical results where they need to go.
```

\begin{singlespace}
\begin{abstract}
Productivity of lentic ecosystems has been well studied and it is widely accepted that as nutrient inputs increase,  productivity increases and lakes transition from low trophic state (e.g. oligotrophic) to higher trophic states  (e.g. eutrophic). These broad trophic state classifications are good predictors of ecosystem health and ecosystem services/disservices (e.g. recreation, aesthetics, fisheries, and harmful algal blooms). While the relationship between  nutrients and trophic state provides reliable predictions, it requires *in situ* water quality data in order to paramterize the model.  This limits the application of these models to lakes with existing and, more importanly, available water quality data.  To expand our ability to predict in lakes without water quality data, we take advantage of the availability of a large national lakes water quality database, land use/land cover data, lake morphometry data, other universally available data, and modern data mining approaches to build and assess models of lake tropic state that may be more universally applied.  We use random forests and random forest variable selection to identify variables to be used for predicting trophic state and we compare the performance of two models of trophic state (as determined by chlorophyll *a* concentration).   The first model estimates trophic state with *in situ* as well as universally available data and the second model uses universally available data only. For each of these models we used three separate trophic state categories, for a total of six models. Overall accuracy for the *in situ* and universal data models ranged from xx% to xx% and xx, xx, and xx described the most variation in trophic state.  For the universal data only models, Overall accuraccy ranged from xx% to xx% and xx, xx, and xx described the most variation in trophic state. Lastly, it is believed that the presence and abundance of cyanobacteria is strongly associated with trophic state.  To test this we examine the association between estimates of cyanobacteria biovolume and the measured and predicted trophic state.  Expanding these preliminary results to include cyanobacteria taxa indicates that cyanobacteria are significantly more likely to be found in highly eutrophic lakes. These results suggest that predictive models of lake trophic state may be improved with additional information on the landscape surrounding lakes and that those models provide additional information on the presence of potentially harmful cyanobacteria taxa. 
\end{abstract}
\end{singlespace}

@@

##Model 1: 4 Trophic States ~ All Variables
```{r ALL_TS4, echo=FALSE, eval=FALSE, cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
all_ts4<-iterVarSelRF(hkm2014Data[predictors_all],hkm2014Data$TS_CHLA_4,numCore=4,100,
    									ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
											vars.drop.num=1,outStr="all_ts4",time=TRUE)
```

```{r ALL_TS4_rf,echo=FALSE,eval=FALSE,cache=T}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
all_ts4_vars<-unique(unlist(all_ts4[1:100]))
all_ts4_rf<-runRandomForest(hkm2014Data[all_ts4_vars],hkm2014Data$TS_CHLA_4, ntree=10000,importance=TRUE,proximity=TRUE)
```

The selected variables that made up Model 1 were Potassium, Nitrogen:Phosphorus, Total Nitrogen, Total Phosphorus, Total Organic Carbon, Turbidity, Ecoregion, Organic Ions, Dissolved Organic Carbon, and Maximum Lake Depth (Table \ref{tab:VarSel_Model1}). Total accuracy for Model 1 is `r round(classAgreement(all_ts4_rf$confusion[,1:4])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(all_ts4_rf$confusion[,1:4])$kappa,3)` (Table \ref{tab:Confusion_Model1}).

@@

```{r ALL_TS3, echo=FALSE, eval=FALSE, cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
all_ts3<-iterVarSelRF(hkm2014Data[predictors_all],hkm2014Data$TS_CHLA_3,numCore=4,100,
  										ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
											vars.drop.num=1,outStr="all_ts3",time=TRUE)
```

```{r all_TS3_rf,echo=FALSE,eval=FALSE,cache=T}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
all_ts3_vars<-unique(unlist(all_ts3[1:100]))
all_ts3_rf<-runRandomForest(hkm2014Data[all_ts3_vars],hkm2014Data$TS_CHLA_3, ntree=10000,importance=TRUE,proximity=TRUE)
```

@@

##Model 3: 2 Trophic States ~ All Variables

```{r ALL_TS2, echo=FALSE, eval=FALSE, cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
all_ts2<-iterVarSelRF(hkm2014Data[predictors_all],hkm2014Data$TS_CHLA_2,numCore=4,100,
  										ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
											vars.drop.num=1,outStr="all_ts2",time=TRUE)
```

```{r all_TS2_rf,echo=FALSE,eval=FALSE,cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
all_ts2_vars<-unique(unlist(all_ts2[1:100]))
all_ts2_rf<-runRandomForest(hkm2014Data[all_ts2_vars],hkm2014Data$TS_CHLA_2, ntree=10000,importance=TRUE,proximity=TRUE)
```

Total accuracy for Model 3 is `r round(classAgreement(all_ts2_rf$confusion[,1:2])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(all_ts2_rf$confusion[,1:2])$kappa,3)`.

@@

##Model 4: 4 Trophic States ~ GIS Only Variables

```{r GIS_TS4, echo=FALSE, eval=FALSE, cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
gis_ts4<-iterVarSelRF(hkm2014Data[predictors_gis],hkm2014Data$TS_CHLA_4,numCore=4,100,
      								ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
											vars.drop.num=1,outStr="gis_ts4",time=TRUE)
```

```{r GIS_TS4_rf,echo=FALSE,eval=FALSE,cache=T}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
gis_ts4_vars<-unique(unlist(gis_ts4[1:100]))
gis_ts4_rf<-runRandomForest(hkm2014Data[gis_ts4_vars],hkm2014Data$TS_CHLA_4, ntree=10000,importance=TRUE,proximity=TRUE)
```

Total accuracy for Model 4 is `r round(classAgreement(gis_ts4_rf$confusion[,1:4])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(gis_ts4_rf$confusion[,1:4])$kappa,3)`.

@@

##Model 5: 3 Trophic States ~ GIS Only Variables

```{r GIS_TS3, echo=FALSE, eval=FALSE, cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
gis_ts3<-iterVarSelRF(hkm2014Data[predictors_gis],hkm2014Data$TS_CHLA_3,numCore=4,100,
    									ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
											vars.drop.num=1,outStr="gis_ts3",time=TRUE)
```

```{r GIS_TS3_rf,echo=FALSE,eval=FALSE,cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
gis_ts3_vars<-unique(unlist(gis_ts3[1:100]))
gis_ts3_rf<-runRandomForest(hkm2014Data[gis_ts3_vars],hkm2014Data$TS_CHLA_3, ntree=10000,importance=TRUE,proximity=TRUE)
```

Total accuracy for Model 5 is `r round(classAgreement(gis_ts3_rf$confusion[,1:3])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(gis_ts3_rf$confusion[,1:3])$kappa,3)`.

@@

##Model 6: 2 Trophic States ~ GIS Only Variables

```{r GIS_TS2, echo=FALSE, eval=FALSE, cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
gis_ts2<-iterVarSelRF(hkm2014Data[predictors_gis],hkm2014Data$TS_CHLA_2,numCore=4,100,
  										ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
											vars.drop.num=1,outStr="gis_ts2",time=TRUE)
```

```{r GIS_TS2_rf,echo=FALSE,eval=FALSE,cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
gis_ts2_vars<-unique(unlist(gis_ts2[1:100]))
gis_ts2_rf<-runRandomForest(hkm2014Data[gis_ts2_vars],hkm2014Data$TS_CHLA_2, ntree=10000,importance=TRUE,proximity=TRUE)
```

Total accuracy for Model 6 `r round(classAgreement(gis_ts2_rf$confusion[,1:2])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(gis_ts2_rf$confusion[,1:2])$kappa,3)`.

>>>>

---
title: 'Modeling Lake Trophic State: A Data Mining Approach'
author: 
- name: Jeffrey W. Hollister
  affilnum: 1
  email: hollister.jeff@epa.gov
- name: W. Bryan Milstead
  affilnum: 1
- name: Betty J. Kreakie
  affilnum: 1
affiliation:
- affilnum: 1
  affil: US Environmental Protection Agency, Office of Research and Development, National     Health and Environmental Effects Research Laboratory, Atlantic Ecology Division, 27 Tarzwell Drive  Narragansett, RI, 02882, USA
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    template: components/manuscript.latex
  html_document: null
  word_document: null
fontsize: 11pt
capsize: normalsize
csl: components/ecology.csl
documentclass: article
spacing: doublespacing
linenumbers: yes
bibliography: components/LakeTrophicModelling.bib
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Modeling Lake Trophic State: A Data Mining Approach}
-->

@@

```{r analysis , eval=FALSE, include=FALSE, echo=FALSE}
##########################################################################################
#All analysis in here, that way all bits of the paper have access to the final objects
#Tables, figures, and numerical results placed at end of paper.
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
##########################################################################################
#Model 1: All Variables, 4 Trophic State Classes
  #Variable Selction
  all_ts4<-iterVarSelRF(hkm2014Data[predictors_all],hkm2014Data$TS_CHLA_4,numCore=4,100,
        								ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  											vars.drop.num=1,outStr="all_ts4",time=TRUE)
  #Random Forest
  all_ts4_vars<-unique(unlist(all_ts4[1:100]))
  all_ts4_rf<-runRandomForest(hkm2014Data[all_ts4_vars],hkm2014Data$TS_CHLA_4, 
                              ntree=10000,importance=TRUE,proximity=TRUE)
##########################################################################################
#Model 2: All Variables, 3 Trophic State Classes
  #Variable Selction
  all_ts3<-iterVarSelRF(hkm2014Data[predictors_all],hkm2014Data$TS_CHLA_3,numCore=4,100,
      									ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  											vars.drop.num=1,outStr="all_ts3",time=TRUE)
  #Random Forest
  all_ts3_vars<-unique(unlist(all_ts3[1:100]))
  all_ts3_rf<-runRandomForest(hkm2014Data[all_ts3_vars],hkm2014Data$TS_CHLA_3, 
                              ntree=10000,importance=TRUE,proximity=TRUE)
##########################################################################################
#Model 3: All Variables, 2 Trophic State Classes
  #Variable Selction
  all_ts2<-iterVarSelRF(hkm2014Data[predictors_all],hkm2014Data$TS_CHLA_2,numCore=4,100,
      									ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  											vars.drop.num=1,outStr="all_ts2",time=TRUE)

  #Random Forest
  all_ts2_vars<-unique(unlist(all_ts2[1:100]))
  all_ts2_rf<-runRandomForest(hkm2014Data[all_ts2_vars],hkm2014Data$TS_CHLA_2, 
                              ntree=10000,importance=TRUE,proximity=TRUE)
##########################################################################################
#Model 4: GIS Only Variables, 4 Trophic State Classes
  #Variable Selction
  gis_ts4<-iterVarSelRF(hkm2014Data[predictors_gis],hkm2014Data$TS_CHLA_4,numCore=4,100,
          							ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  											vars.drop.num=1,outStr="gis_ts4",time=TRUE)

  #Random Forest
  gis_ts4_vars<-unique(unlist(gis_ts4[1:100]))
  gis_ts4_rf<-runRandomForest(hkm2014Data[gis_ts4_vars],hkm2014Data$TS_CHLA_4, 
                              ntree=10000,importance=TRUE,proximity=TRUE)
##########################################################################################
#Model 5: GIS Only Variables, 3 Trophic State Classes
  #Variable Selction
  gis_ts3<-iterVarSelRF(hkm2014Data[predictors_gis],hkm2014Data$TS_CHLA_3,numCore=4,100,
        								ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  											vars.drop.num=1,outStr="gis_ts3",time=TRUE)

  #Random Forest
  gis_ts3_vars<-unique(unlist(gis_ts3[1:100]))
  gis_ts3_rf<-runRandomForest(hkm2014Data[gis_ts3_vars],hkm2014Data$TS_CHLA_3, 
                              ntree=10000,importance=TRUE,proximity=TRUE)
##########################################################################################
#Model 6: GIS Only Variables, 2 Trophic State Classes
  #Variable Selction
  gis_ts2<-iterVarSelRF(hkm2014Data[predictors_gis],hkm2014Data$TS_CHLA_2,numCore=4,100,
      									ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  											vars.drop.num=1,outStr="gis_ts2",time=TRUE)
  #Random Forest
  gis_ts2_vars<-unique(unlist(gis_ts2[1:100]))
  gis_ts2_rf<-runRandomForest(hkm2014Data[gis_ts2_vars],hkm2014Data$TS_CHLA_2, 
                              ntree=10000,importance=TRUE,proximity=TRUE)

```

```{r summaryStats, echo=FALSE, cache=TRUE}
acc_results<-data.frame(model_number=c("Model 1","Model 2","Model 3","Model 4","Model 5","Model 6"),
                        total_accuracy=c(classAgreement(all_ts4_rf$confusion[,1:4])$diag,
                                         classAgreement(all_ts3_rf$confusion[,1:3])$diag,
                                         classAgreement(all_ts2_rf$confusion[,1:2])$diag,
                                         classAgreement(gis_ts4_rf$confusion[,1:4])$diag,
                                         classAgreement(gis_ts3_rf$confusion[,1:3])$diag,
                                         classAgreement(gis_ts2_rf$confusion[,1:2])$diag),
                        kappa_coeff=c(classAgreement(all_ts4_rf$confusion[,1:4])$kappa,
                                         classAgreement(all_ts3_rf$confusion[,1:3])$kappa,
                                         classAgreement(all_ts2_rf$confusion[,1:2])$kappa,
                                         classAgreement(gis_ts4_rf$confusion[,1:4])$kappa,
                                         classAgreement(gis_ts3_rf$confusion[,1:3])$kappa,
                                         classAgreement(gis_ts2_rf$confusion[,1:2])$kappa))

```

\singlespace
\hrule

**Abstract**

Productivity of lentic ecosystems has been well studied and it is widely accepted that as nutrient inputs increase,  productivity increases and lakes transition from low trophic state (e.g. oligotrophic) to higher trophic states  (e.g. eutrophic). These broad trophic state classifications are good predictors of ecosystem health and ecosystem services/disservices (e.g. recreation, aesthetics, fisheries, and harmful algal blooms). While the relationship between  nutrients and trophic state provides reliable predictions, it requires *in situ* water quality data in order to paramterize the model.  This limits the application of these models to lakes with existing and, more importanly, available water quality data.  To expand our ability to predict in lakes without water quality data, we take advantage of the availability of a large national lakes water quality database, land use/land cover data, lake morphometry data, other universally available data, and modern data mining approaches to build and assess models of lake tropic state that may be more universally applied.  We use random forests and random forest variable selection to identify variables to be used for predicting trophic state and we compare the performance of two models of trophic state (as determined by chlorophyll *a* concentration).   The first model estimates trophic state with *in situ* as well as universally available data and the second model uses universally available data only. For each of these models we used three separate trophic state categories, for a total of six models. Overall accuracy for the *in situ* and universal data models ranged from `r round(min(acc_results[1:3,2]),3)`% to `r round(max(acc_results[1:3,2]),3)`% and xx, xx, and xx described the most variation in trophic state.  For the universal data only models, Overall accuraccy ranged from `r round(min(acc_results[4:6,2]),3)`% to `r round(max(acc_results[4:6,2]),3)`% and xx, xx, and xx described the most variation in trophic state. Lastly, it is believed that the presence and abundance of cyanobacteria is strongly associated with trophic state.  To test this we examine the association between estimates of cyanobacteria biovolume and the measured and predicted trophic state.  Expanding these preliminary results to include cyanobacteria taxa indicates that cyanobacteria are significantly more likely to be found in highly eutrophic lakes. These results suggest that predictive models of lake trophic state may be improved with additional information on the landscape surrounding lakes and that those models provide additional information on the presence of potentially harmful cyanobacteria taxa.

\hrule
\doublespace

@@

##Model 1: 4 Trophic States ~ All Variables
The selected variables that made up Model 1 were Potassium, Nitrogen:Phosphorus, Total Nitrogen, Total Phosphorus, Total Organic Carbon, Turbidity, Ecoregion, Organic Ions, Dissolved Organic Carbon, and Maximum Lake Depth (Table \ref{tab:VarSel_Model1}). Total accuracy for Model 1 is `r round(classAgreement(all_ts4_rf$confusion[,1:4])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(all_ts4_rf$confusion[,1:4])$kappa,3)` (Table \ref{tab:Confusion_Model1}).

@@

@@

##Model 3: 2 Trophic States ~ All Variables
Total accuracy for Model 3 is `r round(classAgreement(all_ts2_rf$confusion[,1:2])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(all_ts2_rf$confusion[,1:2])$kappa,3)`.

@@

##Model 4: 4 Trophic States ~ GIS Only Variables
Total accuracy for Model 4 is `r round(classAgreement(gis_ts4_rf$confusion[,1:4])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(gis_ts4_rf$confusion[,1:4])$kappa,3)`.

@@

##Model 5: 3 Trophic States ~ GIS Only Variables
Total accuracy for Model 5 is `r round(classAgreement(gis_ts3_rf$confusion[,1:3])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(gis_ts3_rf$confusion[,1:3])$kappa,3)`.

@@

##Model 6: 2 Trophic States ~ GIS Only Variables
Total accuracy for Model 6 `r round(classAgreement(gis_ts2_rf$confusion[,1:2])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(gis_ts2_rf$confusion[,1:2])$kappa,3)`.
